---
title: "Week 4"
author: "Aditya"
date: "6/20/2020"
output: html_document
---

# Fixing character variables
```{r}

if(!file.exists("cameras.csv")){
  fileUrl <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.csv?accessType=DOWNLOAD"
  download.file(fileUrl,destfile="cameras.csv",method="curl")
}

cameraData <- read.csv("cameras.csv")
names(cameraData)

```

As we can see here, the column names here, sometimes may give us error while naming. Because of the
upper and lower case naming. So what we can do is, make all the names lowercase

```{r}

tolower(names(cameraData))

```

We can also use strsplit() to split the variables.

```{r}

splitNames = strsplit(names(cameraData), '\\.' )
splitNames[5]
splitNames[6]

```

## Quick variables
```{r}

my_list = list(names = c('hakuna','matata','gg boy') , numbers = 1:3, matrix = matrix(1:25,ncol=5) )


my_list$names
cat('\n\n')
my_list$numbers
cat('\n\n')
my_list$matrix

```

## Removing periods from dataframes
```{r}

myf = function(x){x[1]}
sapply(splitNames, myf)

```

```{r}

fileUrl1 <- "https://dl.dropboxusercontent.com/u/7710864/data/reviews-apr29.csv"
fileUrl2 <- "https://dl.dropboxusercontent.com/u/7710864/data/solutions-apr29.csv"
if(!file.exists('reviews.csv')) download.file(fileUrl1,destfile="reviews.csv",method="curl")
if(!file.exists('solutions.csv')) download.file(fileUrl2,destfile="solutions.csv",method="curl")
reviews <- read.csv("reviews.csv"); solutions <- read.csv("solutions.csv")
head(reviews,2)

```

## Replacing characters
```{r}

sub('_','',names(reviews))

```

To replace multiple occurances, use gsub, sub ony replces first occurance
```{r}

my_text = 'hello_world_hakuna_matata'

sub('_','',my_text)
cat('\n\n')
gsub('_','',my_text)


```

## Searching for specific variables, among the variables

```{r}

grep('Alameda',cameraData$intersection)

```

So, let's check now
```{r}

cameraData$intersection[c(65,69,79)]

```

Even we can check, how many alamanda are in there
```{r}

table(grepl('Alameda',cameraData$intersection))
# Non Alameda Data
cameraData2 = cameraData[!grepl('Alameda',cameraData$intersection),]

```

We can actually get the value itself, instead of it's index
```{r}

grep('Alameda',cameraData$intersection,value = T)

```

If Data is not present, it'll simply output zero

```{r}

grep('Aditya',cameraData$intersection)
length(grep('Aditya',cameraData$intersection))

```

## Stringr library
```{r}

library(stringr)
nchar('how many characters are here? bitch')
substr('Aditya sangma',1,7)
paste('hakuna','matata','joto','gula','iccha',sep = '-')

```

Trim extra spaces with str_trim() at the beggining and end
```{r}

str_trim('    Hakuna                 Matata.      It means no worries               ')

```

# Regular expressions
* abc sign will match exactly abc. These means literals. Exact matching
  * abc:
    * abc
    * abc are the first three laters
    * sub name is abc
* ^ sign stands for start of the line:
  * ^I am:(This will match all the sentences starting with I am)
    * I am a bot
    * I am sexy
    * I am mogoda gorob
* $ signs stands for end of the line:
  * kutta$: (this will match all strings ending with kutta)
    * Tui ekta kutta
    * madarchor, kutta
    * eita to ekta bacca kutta
* [Bb][Uu][Sh][Hh]
* [] will match any of the letters inside it
  * the upper one will match all the versions of bush
* Any signs can be used together to form regular expression
  * ^[Ii] am: (This will match all the strings with 'I am' or 'i am' )
    * I am sexy
    * i am mogoda gorob
  * [jJ][uU]$: (This will match all string wnding with all versions of `ju')
    * Nomoskar da, Ju Ju
    * hoda furele hobede ju ju
* [0-9] the sign '-' is used for range
  * [0-9][a-zA-Z]: (This will match anything which starts with a digit and ends with an alphabet)
    * 9am
    * 1pm
    * 9Taka
* [^?.]$ will select end of line ending with anything but period or question mark
* . sign used to refer any character
  * 9.11:
    * 9-11
    * 9a11
    * 9!11
* | sign is used to concat regular expressions. This means eithr this or that
  * hakuna|matata: (This will match hakuna or matata)
    * hakuna matata it means no worries
    * hey hakuna
    * hey matata
    * tomma sedar hakuna
  * earthquake|harikane|volcano|flood|tornado
* Examples:
  * \^[Gg]ood|[Bb]ad:(Starting with 'good', or anywhere in the line 'bad')
    * Good morning
    * Goo boy, bad boy
    * good looking
    * kutta bad er mayre ch**di
    * Hakuna matata, this is not bad
  * ^([Gg]ood|[Bb]ad) Same as above. But it's good to add this. To maintain constrains
* ? mark means the indicated expression is optional.
* for special characters like . \ ( * to match, escape character \ must be added
* It's important to remember that, ' ' spaces, tabs anything put in the regular expression wiil be 
put for matching. So be careful when writing the regular expressions.
* [Gg]eorge( [Ww]\. )?[Bb]ush will match all the versions of George W.Bush
  * George W. Bush
  * george Bush
  * George bush
  * george w. Bush
* \* means repeating any number of times even 0 times
* a*:
  * a
  * 
  * aa
  * aaaaaaaaaaaaaaaa
* *.\** :
  * aadjkflkajdfl;k
  * aditya
  * lesson
  *
  * memesa
  * nancy
  * tomma ghile
  * apan
* + means at least one of the item
* a+ wont match '' but a* will
* [a-zA-Z_]+([0-9]\*|[a-zA-Z_])* this is used for variables name declaration:
  * abc9
  * abc
  * ____
  * _19
  * a_b
* {} sign is used for ranging, how many times this need to be repeated
  * (abc){1,5} means match between 1 to 5
* m,n means between m and n
* m means exactly m times
* m, means at leas m times
* Regular expression like escape numbers \1 \2 can be used to repeat patterns. Regular expression 
compiler remembers these expressions. For example:
  *  +[a-zA-Z]+ +\1 + : (at least one space followed by at least one character followed by at least
  one space and then, exactly the same literal matched by the first expression followed by at least
  one space and so on....)
    * blah blah blah
    * time for night night witcher
    * sugorot glasaba hijok hojok guri dim
    * indi ele, veret veret guri pidim
* \* is greedy. It matches the lognest possible match.
  * ^s(.*)s: starting with s ending with s
    * settting up all the sets
    * sakura tomma ghile, mogdar bitchs
* to avoid it, we can add question mark
  * ^s(.*)?s$



## Dates

```{r}

d1 = date()
d1

```
It's actually a string of characters
```{r}

class(d1)

```

```{r}

d2 = Sys.Date()
d2

```

```{r}

class(d2)

```

Formatting Data is possible using format
```{r}

format(d2, '%a %b %d')
format(d2,'%A %B %d')

```
%d day as number, %a unabbreviated weekday, %A abbreviated weekday, %m month (00-12), %b unabbreviated
month name, %B abbreviated month name, %y 2 digit year, %Y 4 digit Year


Character vector can be turned into dates using as.Date() function. We need to teel the function, the format dates are in
```{r}

x = c('1jan1960', '29feb1996', '21jun2020' ); z = as.Date(x,'%d %b %Y')
z

```

Date can be used with arithmatic expressions
```{r}

Sys.Date() - z[2]

```

Different type of functions

```{r}

months(d2)
weekdays(d2)
julian(d2)

```
Julian is a special type of function that tells, exactly how many days since the origin. And it'll also tell us the origin.

## Lubridate
This library makes easy to work with dates
```{r}

library(lubridate)
ymd('19960229')

```

```{r}

mdy('06212020')

```
```{r}

ymd_hms('2011-08-03, 10:55:33')

```

We can also set the timezone
```{r}

ymd_hms('2011-08-03, 10:55:33',tz='Pacific/Auckland')

```

```{r}

Sys.timezone()

```

In this library some functions have different syntaxes
```{r}

wday(Sys.Date())

```
```{r}

wday(Sys.Date(),label = T)

```



